Classless Java: Tuning Java Interfaces

### 2015-11-25 paper/sections/RelatedWork.tex
Revised Related work (in progress)

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  It provides
expressive power, but it is difficult to model and implement, and can create
programs that are hard to reason about. These difficulties include the famous
diamond (fork-join) problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.
%and the yo-yo problem~\cite{taenzer1989problems}.  
To conciliate the need for expressive power and the need for simplicity, lots of
models have been proposed in the past few years, including C++ virtual
inheritance, mixins~\cite{bracha90mixin} and traits~\cite{scharli03traits}.
They provide novel programming architecture models in the OO paradigm. In terms
of restrictions set on these models, C++ virtual inheritance aims at a more
general model, mixins added some restrictions on the model, and trait model is
the most restricted one (excluding states, instantiation, etc).

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance. Virtual
inheritance in C++ provides another solution to multiple inheritance (especially
the diamond problem by keeping only one copy of the base
class)~\cite{ellis1990annotated}, however suffers from object initialization
problem as pointed out by Malayeri et al.~\cite{malayeri2009cz}. It bypasses all
constructor calls to virtual superclasses, which would potentially cause serious
semantic error.

\paragraph{Mixins.}
Mixins are a more restricted model than C++ approach. It allows to name
components that can be applied to various classes as reusable functionality
units. However, they suffer from linearisation: the order of mixin application
is relevant in often subtitle and undesired ways. This hinders their usability
and the ability of resolving conflicts: the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in all cases and
restricts the flexibility of mixin composition. To fight those limitations, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raised the complexity of the approach, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala solution adopts linearised mixins,
(but misleadingly called traits in the language) avoids object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition.
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@

\noindent Our approach does not involve linearised mixin problem, in that the semantics of
Java \textbf{extends} clause is unordered and symmetric. Also,our approach
supports proper constructor mechanism.

\paragraph{Traits.}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a strong
line between units of reuse (traits) and object factories (classes). In this
model, traits as units of reusable code, contain only methods as reusable
functionalities. Thus, no state/state initialization is considered. Classes act
as object factories, requiring functionalities from multiple traits.  Traits
offer a trait algebra with operations like sum, alias and exclusion, provided
for explicit conflict resolution.  Concluding, (pure) traits do not allow state
and they do not offers any reuse instrument to ensure that state is coherently
initialized when finally defined in classes.  Traits can not be
instantiated. This model requires two concepts (traits and classes) to coexist
and cooperate.  Some authors see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model. Our approach takes interfaces as
units of reuse, and meanwhile generates factory method for instantiation.

@@

\subsection{Multiple Inheritance in Java}
Since Java 8 default methods are introduced, concrete method implementation are
allowed to be defined (via the \textbf{default} keyword) inside
interfaces. 
% Since Java supports implementation of multiple interfaces (instead
% of extension of a single class), 
The introduction of default methods opens the
gate for various flavours of multiple inheritance in Java, using interfaces.
Former work by Bono.et. al.~\cite{bono14}. provides details on mimicking traits
through interfaces.

There are proposals for extending Java (before Java8) with traits. For example,
FeatherTrait Java (FTJ)~\cite{Liquori08ftj} by Liquori et al. extends the
calculus of Featherweight Java (FJ)~\cite{Igarashi01FJ} with statically-typed
traits, adding trait-based inheritance in Java.  Except for few, mostly
syntactic details, their work can be emulated/rephrased with Java8 interface.

\input{sections/Compare.tex}

>>>>

\subsection{Multiple Inheritance in Object Oriented Languages}
Many authors have argued in favour or against multiple inheritance.  Multiple inheritance provides
expressive power, but it is difficult to model and implement, and can create
programs that are hard to reason about. These difficulties include the famous
diamond (fork-join) problem~\cite{bracha90mixin,Sak89dis}, conflicting methods, etc.
%and the yo-yo problem~\cite{taenzer1989problems}.  
To conciliate the need for expressive power and the need for simplicity, many
models have been proposed in over the years, including C++ virtual
inheritance, mixins~\cite{bracha90mixin} and traits~\cite{scharli03traits}.
They provide novel programming architecture models in the OO paradigm. In terms
of restrictions set on these models, C++ virtual inheritance aims at a more
general model, mixins added some restrictions on the model, and trait model is
the most restricted one (excluding states, instantiation, etc).
\bruno{How about Malayeri and Aldrich's paper? shouldn't that be discussed?}

\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance. Virtual
inheritance in C++ provides another solution to multiple inheritance (especially
the diamond problem by keeping only one copy of the base
class)~\cite{ellis1990annotated}, however suffers from object initialization
problem as pointed out by Malayeri et al.~\cite{malayeri2009cz}. It bypasses all
constructor calls to virtual superclasses, which would potentially cause serious
semantic errors. \bruno{What happens in our approach for the same case?}

\paragraph{Mixins.}
Mixins are a more restricted model than the C++ approach. Mixins allow to name
components that can be applied to various classes as reusable functionality
units. However, they suffer from linearisation: the order of mixin application
is relevant in often subtle and undesired ways. This hinders their usability
and the ability of resolving conflicts: the linearisation (total ordering) of
mixin inheritance cannot provide a satisfactory resolution in all cases and
restricts the flexibility of mixin composition. To fight those limitations, an
algebra of mixin operators is introduced~\cite{ancona2002calculus}, but this
raised the complexity of the approach, especially when constructors and fields
are considered~\cite{marco09FJigsaw}. Scala traits are in fact more like linearised mixins.
Scala avoids the object initialization
problem by disallowing constructor parameters, causing no ambiguity in cases
such as diamond problem. However this approach has limited expressiveness, and
suffers from all the problems of linearised mixin composition. Other languages, such as 
Python, also use linearized mixins.
\begin{comment}
Python also offers multiple inheritance via linearised mixins. Indeed in python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux): pass@ 
\end{comment}
\noindent Java interfaces and default methods does not use
linearisation: the semantics of Java \textbf{extends} clause in
interfaces is unordered and symmetric.

\bruno{rrevise latter?}
However, in pure Java, there is no mechanism for creating objects in
interfaces. Also, our approach supports proper constructor mechanism.

\paragraph{Traits and Java's default methods}
Simplifying the mixins approach, traits~\cite{scharli03traits} draw a
strong line between units of reuse (traits) and object factories
(classes). In this model, traits as units of reusable code, contain
only methods as reusable entities. Thus, no state and state
initialization are considered. Classes act as object factories,
requiring functionalities from multiple traits. Java 8 interfaces with
default methods are closely related to traits: concrete method
implementation are allowed (via the \textbf{default} keyword) inside
interfaces. The introduction of default methods opens the gate for
various flavours of multiple inheritance in Java, using interfaces.
Traits offer a trait algebra with operations like sum, alias and exclusion, provided
for explicit conflict resolution.  
Former work by Bono et al.~\cite{bono14}. provides details on
mimicking the trait algebra through Java 8 interfaces. 
We briefly recall the main points of their encoding;
however we propose a different representation of \textbf{exclusion}.
The author of ~\cite{bono14} agree that our revised version is
cleaner, typesafe and more direct.\bruno{do we want do say this?}

\newcommand\shortItem{\vspace{-1ex}}
\begin{itemize}
\item \textbf{Symmetric sum} can be obtained by simple multiple inheritance between interfaces.
\shortItem
    \begin{lstlisting}
    interface A { int x(); }    interface B { int y(); }    interface C extends A, B {}
    \end{lstlisting}
\shortItem
\item \textbf{Overriding} a conflict is obtained by specifying which super interface take precedence.
\shortItem
    \begin{lstlisting}
    interface A { default int m() {return 1;} } 
    interface B { default int m() {return 2;} }
    interface C extends A, B { default int m() {return B.super.m();} }
    \end{lstlisting}
\shortItem
\item \textbf{Alias} is creating  a new method delegating to the existing super interface.
\shortItem
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { default int k() {return A.super.m();} }
    \end{lstlisting}
\shortItem

\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces.
\shortItem
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\shortItem
\end{itemize}

There are also proposals for extending Java (before Java8) with
traits. For example, FeatherTrait Java (FTJ)~\cite{Liquori08ftj} by
Liquori et al. extends the calculus of Featherweight Java
(FJ)~\cite{Igarashi01FJ} with statically-typed traits, adding
trait-based inheritance in Java.  Except for few, mostly syntactic
details, their work can be emulated/rephrased with Java8 interfaces.
There are also extensions to the original trait model, which have operations 
that default methods and interfaces cannot model, such as method renaming
(as in [Reppy2006]), which breaks structural subtyping.

\paragraph{Traits vs Object Interfaces}
We consider object interfaces to be an alternative to traits or mixins.
In the trait model two concepts (traits and classes) coexist and
cooperate. Some authors~\bruno{who?} see this as good language design
fostering good software development by helping programmers to think
about the structure of their programs.  However, other authors see the
need of two concepts and the absence of state as drawbacks of this
model~\cite{malayeri}.  Object interfaces are units of reuse, and at
the same time they provide factory methods for instantiation and
support state.  Our approach promotes the use of interfaces instead of
classes, in order to rely on the modular composition offered
by interfaces. Since Java was designed for classes, a direct class-less
programming style is verbose and feels unnatural. However, annotation
driven code generation is enough to overcome this difficulty, and the
resulting programming style encourages modularity, composability and
reusability, by keeping a strong object oriented feel. In that sense,
we promote object interfaces as being both units of reusable code and
object factories. Our practical experience is that, in Java,
separating the two notions leads to alot of boilerplate code, and is
quite limiting when multiple inheritance with state is required. The
use of abstract state operations avoids the key difficulties
associated with multiple inheritance and state, while still being quite expressive. 
Moreover the ability to support constructors 
adds additional expressivity, which is not available in approaches such 
as Scala's traits/mixins.

%Nevertheless, we
%believe that in other languages the separation can be more practical
%than in Java, and we do not necessarally advocate that merging the two
%notions is a better approach in general.

%There are other limitations of our current approach, but they may be addressed
%in future work (see Section~\ref{sec:futurework}).

\begin{comment}
Besides, we support more features than the original trait model:
\begin{itemize}
\item We provide \texttt{of} methods for the annotated interfaces. During annotation processing time, the ``fields'' inside an interface are observed and a static method \texttt{of} is automatically injected to the interface with its arguments correspondingly. Such a method is a replacement to the constructors in original traits, making instantiation more convenient to use.
That is, in our approach there are only interfaces, our model requires a single concept, while the trait model requires traits \emph{and} classes.

\item We provide \texttt{with-} methods as auxiliary constructors. A \texttt{with-} method is generated for each field, just like record update, returning the new object with that field updated
%. A \texttt{clone} method is generated for the interface, returning a copy of the current object.
Furthermore, we do automatic return type refinement for these kind of methods. This feature is comparatively useful in big examples, making operations and behaviours more flexible.%, which we will demonstrate later.
\item We provide two options for generating setters. There are two kind of setters which are commonly used, namely \textit{void setters} and \textit{fluent setters}. The only difference is that a fluent setter returns the object itself after setting, thus supporting a pipeline of such operations. The generation depends on which type of setter is declared in the interface by users.
\end{itemize}
\end{comment}

@@

% \input{sections/Compare.tex}
